---
title: "First tests in R"
author:
  - Sophie C. Schmidt:
      email: s.c.schmidt@uni-koeln.de
      correspondence: true
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  md_document:
    variant: markdown_github
always_allow_html: true
bibliography: references.bib
csl: RGK_archaeology_DGUF.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      include = TRUE)

```



# Tests!

## binomial test

This tests checks whether two nominal values, that are exclusive to each other, follow a theoretical distribution.

Let's imagine we've got a graveyard with 200 graves and we were able to determine exactly whether they were male or female. So there are 75 women and 125 men in our dataset.

These numbers make us think and we develop a hypothesis: There were as many women and men in the society that used the graveyard. We therefore believe that less women were entombed in this graveyard than men. But of course our excavation is only a sample, therefore we need to check whether that's statistically plausible:

From this we can develop the statistical hypothesis:

- H0 =  There is no difference between the graves of men and women in the population. The difference in our sample is by chance.
- H1 = non-directed: There are differences between the number of men and women interred in the graveyard. directed: Men were more often entombed than women

Der Binomialtest ist in R base umgesetzt und sehr leicht ausführbar. The general syntax is  `binom.test(nsuccesses, ntrials, p for successes, alternative="greater / less / two.sided")`. To translate this in our example:

nsuccesses would be men interred, ntrials are the number of graves we excavated, p is the statistical chance it should be a man (50%) and alternative (hypothesis) would be "greater" if we say number of successes is greater than expected, "less" if the number of successes are less than expected and "two.sided" if we do a non-directed test.

```{r binomial Bsp}
binom.test(125, 200, 0.5, alternative="greater")

# the other way round is just as possible:
binom.test(75, 200, 0.5, alternative = "less") # now the 75 Frauengräber are "successes" and therefore the alternative hypothesis needs to be turned around to "less"

# without a direction in the alternative hypothesis
binom.test(75, 200, 0.5, alternative = "two.sided")
```
Easy no?

Let's do this with a real archaeological data set. Let's just take the same question with the burial ground Ernest Witte in Texas:

```{r data sets}
library(archdata)
data(EWBurials)
```
Maybe first have a look at what the data set consists of using `?EWburials`.

What we are going to do is write small functions inside our test-function. 
We first ask, how many burials are in the dataset alltogether (ntrials). For this we can use `nrow`:

```{r}
nrow(EWBurials)
```

Now we need to now, how many of those are of men? So we do a subset of `EWBurials`, where `EWBurials$Sex` equals "Male":

```{r}
nrow(EWBurials[EWBurials$Sex == "Male",])
```

don't forget the comma at the end: Remeber, using square brackets [,] enabled us to choose certain values from a table by selecting rows and columns. We can also add a condition. And we want to filter the rows for where the column Sex gives "Male", so we put this condition in the "rows" part inside the square brackets. But the columns part must "still be there", even if it stays empty.

Now, we don't want to copy and paste the values we learned from the `nrow()` calls, that's bad practice. We will just put the code inside the function:

```{r EWBurials binomial}

binom.test(nrow(EWBurials[EWBurials$Sex == "Male",]), 
           nrow(EWBurials), 
           0.5, alternative = "two.sided")
```

How can we interprete the result?
The p-value is 1! 

So there is a 100% chance that we are WRONG if we accept the alternative hypothesis... soo... yes, here we have a 50-50 distribution of men and women. No surprise for those who looked at the numbers before.

But that's how easy the binomial test is!
